Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2019-12-23 analysis/paper/paper.Rmd
work on MANOVA table

The elliptic Fourier coefficients are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variances. Figure \@ref(fig:contribution-PCs) shows the shape variation associated to the first three principal components. From the negative to the positive scores, the PC1 represents the height of the vessels from tall to short and the roundness of the body from round to oval shaped. The PC2 relates to the neck and mouth constriction from narrow to wide. The PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of flare of neck from curved to straight shape. The first two components account for most of the variance in relation to three phases were represented in Figure \@ref(fig:first-two-PCs-byplot). The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European period that shows some extent of similarity and consistency. However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European period compared to those in Chinese period along both PC1 and PC2 axes. In other words, there is a decreasing in shape variance in the Chinese period that shares the features of shorter height and narrower mouth.

```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes. The rows from the top to the bottom represent the first to third principal component respectively. Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
# the contributions of each PC to the shapes
pc_contrib <- PCcontrib(op.p, nax = 1:3) 
dev.off()
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  theme_minimal(base_size = 6)

@@

```{r MANOVA}
# MANOVA: Multivariate Analysis of (co)variace
# We can test for a difference in the distribution of PC scores by calculating a pairwise combination between every levels of a fac.

manv <-   MANOVA_PW(op.p, "phase")
# manv$summary

#----------------------------------------------------
# GAMMS from https://www-nature-com.offcampus.lib.washington.edu/articles/s41598-018-20122-9

@@

ggplot(pottery_data,
       aes(phase,
           as.numeric(Thick1_Neck))) +
  geom_boxplot()

@@

library(brms)
m2 <- brm(bf(shape_pc1 ~ s(thick_body1)),
          data = gams_input, family = gaussian(), cores = 4, seed = 17,
          iter = 4000, warmup = 1000, thin = 10, refresh = 0,
          control = list(adapt_delta = 0.99))
summary(m2) # so many zeros... not very promising 
# plot marginal effects for each predictor
plot(marginal_effects(m2), ask = FALSE)
loo(m2)
pp_check(m2)
```

@@

Multivariate analysis of variance (MANOVA) is used to test the difference in three shape varaibles from three phases. The result shows significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01). The distributions of the first three PCs for each vessel suggest variations in shape standardization by period. The first PC, the hight of the vessels, shows more variations in the pre-Euro period compared to the Chinese period. Standardization appears strong in the pre-Euro period but weak in the Chinese period. P-values for a modified signed-likelihood ratio test of equality of CVs of PC1 & PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.

>>>>

The elliptic Fourier coefficients are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variances. Figure \@ref(fig:contribution-PCs) shows the shape variation associated to the first three principal components. From the negative to the positive scores, the PC1 represents the height of the vessels from tall to short and the roundness of the body from round to oval shaped. The PC2 relates to the neck and mouth constriction from narrow to wide. The PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of flare of neck from curved to straight shape. The first two components account for most of the variance in relation to three phases were represented in Figure \@ref(fig:first-two-PCs-byplot). The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European period that might show some extent of similarity. However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European period compared to those in Chinese period along both PC1 and PC2 axes. In other words, there is a decreasing in shape variance in the Chinese period that shares the features of shorter height and narrower mouth.

```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes. The rows from the top to the bottom represent the first to third principal component respectively. Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
# the contributions of each PC to the shapes
pc_contrib <- PCcontrib(op.p, nax = 1:3) 
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  theme_minimal(base_size = 6)

@@

To test differences in the distribution of shape variables indicated by the PC scores, we use multivariate analysis of variance (MANOVA) to compare a pairwise combination between every sample means across three phases. The results in Table \@ref(tab:MANOVA-table) show significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01), which corresponds to the previous observation of the different distribution in shapes in Chinese period according to PCA byplot. In addition, although there is a considerable overlap of shape variables between Pre-European and Post-European, their PC scores are significantly different to each other. There is not much differences between the shapes in Post-European and Chinese contact period.

```{r MANOVA-table}
# MANOVA: Multivariate Analysis of (co)variace
# We can test for a difference in the distribution of PC scores by calculating a pairwise combination between every levels of a fac.

manv <- MANOVA_PW(op.p, "phase")
su_manv <- manv$summary

knitr::kable(su_manv,
             caption = "The statistics summary of MANOVA test for the PC scores")
```

```{r}
#----------------------------------------------------
# GAMMS from https://www-nature-com.offcampus.lib.washington.edu/articles/s41598-018-20122-9

@@

# metric data
ggplot(pottery_data,
       aes(phase,
           as.numeric(Thick1_Neck))) +
  geom_boxplot()

@@

# Bayesian Regression Models using 'Stan'
library(brms)
m2 <- brm(bf(shape_pc1 ~ s(thick_body1)),
          data = gams_input, family = gaussian(), cores = 4, seed = 17,
          iter = 4000, warmup = 1000, thin = 10, refresh = 0,
          control = list(adapt_delta = 0.99))
summary(m2) # so many zeros... not very promising 
# plot marginal effects for each predictor
plot(marginal_effects(m2), ask = FALSE)
loo(m2)
pp_check(m2)
```

@@

The distributions of the first three PCs for each vessel suggest variations in shape standardization by period. The first PC, the hight of the vessels, shows more variations in the pre-Euro period compared to the Chinese period. Standardization appears strong in the pre-Euro period but weak in the Chinese period. P-values for a modified signed-likelihood ratio test of equality of CVs of PC1 & PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
